Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-13 background.tex
revisions up through end of section 3

\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables.png}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

@@

The key pieces of functionality a VMM must implement itself, or have available in another OS component (depending on 
design), are
adding a new page mapping (whether the mapped page contains zeros, file data, or swap data), and removing an existing 
page mapping.
While this initially sounds like relatively modest functionality whose implementation may be complicated by hardware 
subtleties, a number of more serious complexities must be handled.

\todo[inline]{Colin says: we should chat about the paragraph below. We don't want to advertise too many
challenges of VMM verification that we don't actually solve (or rather, we'd push them to the and of the paper
as future work). But some of these I think we could split the difference with, for example by defining
a variant of virtual-pointsto that allowed virtual address aliasing, but not actually verifying any code examples
with it (just showing it's compatible with our approach).
}
For one, virtual addresses may \emph{alias} when more than one is mapped to the same physical memory. 
This can occur when a process requests it (e.g., via \texttt{mmap} or \texttt{shm\_open} on UNIX-style kernels). 
It also occurs automatically in most general-purpose kernels: while for normal programs the OS uses the MMU to isolate 
and abstract physical memory, the kernel's own functionality is often easier to implement if the kernel can directly 
access memory given its physical address, for example when interacting with hardware devices. For this reason, most 
kernels contain an \emph{identity map} of all physical memory in a machine. 
\todo[inline]{We should try defining an identity map on a whiteboard and see what breaks. If it's feasible we should
tweak and get a definition in the paper. If not, we should cut the identity map discussion.}
In some cases this is a literal identity map, 
where for every virtual address $p$, interpreting $p$ as a virtual address instead will map back to $p$ as a physical 
address. In other cases there is an offset involved, where to access physical address $p$ the kernel can access virtual
 address $p+\mathit{offset}$. This identity map saves the kernel the performance cost of having to continually add and
 remove transient mappings simply to briefly access a specific physical memory region. But this coexists with additional
 virtual addresses (those used by most kernel code and data structures), in a form of virtual address aliasing. Note 
that virtual address aliasing is both prevalent in general-purpose kernels, and violates the core assumption of most 
memory models assumed by verified compilers like CompCert, which assume no virtual address aliasing.

\todo[inline]{Colin says: Again, let's chat about the next paragraph. This might be another where we can
split the difference on extension by defining something and proving a $\ast-\ast$ kind of result for sharing
page table fragments, rather than actually proving any new code. I'm less sure about this one, though.
}
Another complicating factor is that portions of page tables can be shared, i.e., trees of page tables can overlap and
become forests. 
One common way this used to happen was for one subtree of page tables to map the kernel if linked in appropriately to 
a higher-level page table, and then to reuse that kernel mapping portion in every process's page tables. This was mostly 
phased out in the wake of Spectre, but remains relevant on correctly-implemented hardware.

These dove-tail with functionality in the OS scheduler, which since it is responsible for dealing with multiple 
address spaces, must keep track of which virtual addresses are valid (and in what way) in which address spaces, 
where some virtual addresses are valid in only a single address space (e.g., a code address for a particular usermode 
process), while others are valid in all address spaces (e.g., kernel data structure pointers). 
The VMM must maintain some of these assumptions on behalf of the rest of the kernel, for example by guaranteeing that 
a certain range of virtual addresses (corresponding to the kernel's code and data) are valid in every address space.

@@

Most modal logics use unary operators on propositions to express that a logical claim $P$ is \emph{contingently} true 
in certain other circumstances, such as in other times~\cite{pnueli1977temporal} or places~\cite{gordon2019modal}. That is, for 
different modal operators $M$, $M(P)$ may mean that $P$ is true in another time or place (which may be a physical place 
like another network node~\cite{murphy2008type,gordon2019modal} or a conceptual place like a person's 
mind~\cite{hintikka1962knowledge}). A unifying concept across any modality is that they behave as applicative functors, 
typically satisfying (directly, or as a derived law):
\[ (P\rightarrow Q) \rightarrow M(P) \rightarrow M(Q)\]
%\todo[inline]{Ismail, I think this means we have pure intro, $\square P \mathrel{-\ast} [r](\square P)$ if I'm using the right symbol for pure assertions }
%\todo[inline,color=red]{The above todo comment isn't quite right.
%Purity is about being able to duplicate. You had defined another typeclass/property-of-assertions that meant it didn't
%care what address space it was in (like physical pointstos). That's orthogonal to purity.
%}
Many modalities, so-called \emph{normal} modalities also possess introduction rules of the form $P\rightarrow M(P)$, 
the classic example being that if $P$ is true, then $P$ is \emph{necessarily} true with the contingency picked up.%($\square P$).

>>>>

\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables.png}
  \todo[inline]{this image got \emph{very} pixelated. is this possible to fix?}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

@@

The key pieces of functionality a VMM must implement itself, or have available in another OS component (depending on 
design), are
adding a new page mapping (whether the mapped page contains zeros, file data, or swap data), and removing an existing 
page mapping.
While this initially sounds like relatively modest functionality whose implementation may be complicated by hardware 
subtleties, even these basic operations are actually quite intrictate.
Most notably, updates to the page tables are performed as writes to memory --- which are themselves subject to address translation.
In the case of changing the mappings for the currently-active set of page tables, the OS kernel is modifying the tables involved in its
own access of the tables.

% \todo[inline]{Colin says: we should chat about the paragraph below. We don't want to advertise too many
% challenges of VMM verification that we don't actually solve (or rather, we'd push them to the and of the paper
% as future work). But some of these I think we could split the difference with, for example by defining
% a variant of virtual-pointsto that allowed virtual address aliasing, but not actually verifying any code examples
% with it (just showing it's compatible with our approach).
% }
% For one, virtual addresses may \emph{alias} when more than one is mapped to the same physical memory. 
% This can occur when a process requests it (e.g., via \texttt{mmap} or \texttt{shm\_open} on UNIX-style kernels). 
% It also occurs automatically in most general-purpose kernels: while for normal programs the OS uses the MMU to isolate 
% and abstract physical memory, the kernel's own functionality is often easier to implement if the kernel can directly 
% access memory given its physical address, for example when interacting with hardware devices. For this reason, most 
% kernels contain an \emph{identity map} of all physical memory in a machine. 
% \todo[inline]{We should try defining an identity map on a whiteboard and see what breaks. If it's feasible we should
% tweak and get a definition in the paper. If not, we should cut the identity map discussion.}
% In some cases this is a literal identity map, 
% where for every virtual address $p$, interpreting $p$ as a virtual address instead will map back to $p$ as a physical 
% address. In other cases there is an offset involved, where to access physical address $p$ the kernel can access virtual
%  address $p+\mathit{offset}$. This identity map saves the kernel the performance cost of having to continually add and
%  remove transient mappings simply to briefly access a specific physical memory region. But this coexists with additional
%  virtual addresses (those used by most kernel code and data structures), in a form of virtual address aliasing. Note 
% that virtual address aliasing is both prevalent in general-purpose kernels, and violates the core assumption of most 
% memory models assumed by verified compilers like CompCert, which assume no virtual address aliasing.

% \todo[inline]{Colin says: Again, let's chat about the next paragraph. This might be another where we can
% split the difference on extension by defining something and proving a $\ast-\ast$ kind of result for sharing
% page table fragments, rather than actually proving any new code. I'm less sure about this one, though.
% }
% Another complicating factor is that portions of page tables can be shared, i.e., trees of page tables can overlap and
% become forests. 
% One common way this used to happen was for one subtree of page tables to map the kernel if linked in appropriately to 
% a higher-level page table, and then to reuse that kernel mapping portion in every process's page tables. This was mostly 
% phased out in the wake of Spectre, but remains relevant on correctly-implemented hardware.

Virtual memory concerns propagate to the OS scheduler, which is responsible for dealing with multiple 
address spaces, so must keep track of which virtual addresses are valid (and in what way) in which address spaces. 
Some virtual addresses are valid in only a single address space (e.g., a code address for a particular usermode 
process), while others are valid in all address spaces (e.g., kernel data structure pointers). 
The VMM must maintain some of these assumptions on behalf of the rest of the kernel, for example by guaranteeing that 
a certain range of virtual addresses (corresponding to the kernel's code and data) are valid in every address space.

@@

Most modal logics use unary operators on propositions to express that a logical claim $P$ is \emph{contingently} true 
in certain other circumstances, such as in other times~\cite{pnueli1977temporal} or places~\cite{gordon2019modal}. That is, for 
different modal operators $M$, $M(P)$ may mean that $P$ is true in another time or place (which may be a physical place 
like another network node~\cite{murphy2008type,gordon2019modal} or a conceptual place like a person's 
mind~\cite{hintikka1962knowledge}). A unifying concept across any modality is that they behave as applicative functors, 
typically satisfying (directly, or as a derived law, depending on the modality):
\[ (P\rightarrow Q) \rightarrow M(P) \rightarrow M(Q)\]
%\todo[inline]{Ismail, I think this means we have pure intro, $\square P \mathrel{-\ast} [r](\square P)$ if I'm using the right symbol for pure assertions }
%\todo[inline,color=red]{The above todo comment isn't quite right.
%Purity is about being able to duplicate. You had defined another typeclass/property-of-assertions that meant it didn't
%care what address space it was in (like physical pointstos). That's orthogonal to purity.
%}
Many modalities, so-called \emph{normal} modalities also possess introduction rules of the form $P\rightarrow M(P)$, 
the classic example being that if $P$ is true, then $P$ is \emph{necessarily} true with the contingency picked up.%($\square P$).

### 2023-04-13 intro.tex
+
revisions up through end of section 3

\section{Introduction}
\label{sec:intro}
\mytodo{Colin's Part}
Virtual memory management lies at the core of modern OS kernel implementation. It is deeply intertwined with most other parts of a typical general-purpose OS kernel design, including scheduling, hardware drivers, and even the filesystem buffer cache. In writing the authoritative reference on the internals of the Solaris kernel, McDougall and Mauro went so far as to claim that ``\emph{the virtual memory sub-system can be considered the core of a Solaris instance, and the implementation of Solaris virtual memory affects just about every other subsystem in the operating system}''~\cite{mcdougall2006solaris}.
This makes rich support for verification the virtual memory management subsystem of an OS kernel critical to the correctness of every other piece of an OS or any software running atop it.

@@

This kind of context-dependent assertion, where a fact may be true in one address space but not others, has a modal flavor. 
We propose tackling the verification of virtual memory subsystems (and kernels more broadly) by adapting ideas from hybrid
modal logic, which can label assertions true under \emph{other, named} circumstances (i.e., in another address space) with a 
modality indexed by a name for that space (in our case, the root of the page tables for an address space). This offers a 
\textit{convenient} and \textit{powerful} way to \emph{modularly}
\begin{itemize}
\item isolate assertions specific to a particular address space,
\item explicitly state when an assertion is true across address spaces,
\item manipulate address spaces from within other address spaces, and
\item reason about change in address spaces.
\end{itemize}
These advantages make this approach to reasoning about virtual memory more flexible than prior program logic techniques~\cite{kolanski08vstte,kolanski09tphols}, 
which were only able to work with a single address space (the current address space on the CPU) because they were unable
to speak directly \emph{within the logic} about other address spaces.

\begin{figure}
   \includegraphics[width=0.5\columnwidth]{architecture.png}
  \caption{Layers of Implementation}
  \todo[inline]{Colin says: I don't think this figure helps readers; it is too abstract}
  \label{fig:architecture}
\end{figure}

We realize and show our reasoning principles, shown in Figure \ref{fig:architecture}
\begin{itemize}
\item We develop these ideas in the form of a logic for working with virtual-address-space-relative assertions,
      implemented as an embedded separation logic within Iris~\cite{jung2018iris} separation logic.
\item We prove the soundness of our \textsf{vProp} logic with respect to a RISC-like fragment of \textsf{AMD64} instructions.
\item We verify simplified versions of several critical virtual-memory-related pieces of OS functionality, 
      including mapping  pages, the assembly code for switching address spaces, and code which manipulates page
      tables for \emph{currently mapped memory} without disturbing mappings.\todo[inline]{Colin says: I have optimistically
      added this last example to the text here, because I think we can probably do it. Leave this TODO
      here until we either do or cut it.}
      The latter in particular goes beyond the capabilities of prior work.
\end{itemize}

>>>>

\section{Introduction}
\label{sec:intro}
Virtual memory management lies at the core of modern OS kernel implementation. It is deeply intertwined with most other parts of a typical general-purpose OS kernel design, including scheduling, hardware drivers, and even the filesystem buffer cache. In writing the authoritative reference on the internals of the Solaris kernel, McDougall and Mauro went so far as to claim that ``\emph{the virtual memory sub-system can be considered the core of a Solaris instance, and the implementation of Solaris virtual memory affects just about every other subsystem in the operating system}''~\cite{mcdougall2006solaris}.
This makes rich support for verification the virtual memory management subsystem of an OS kernel critical to the correctness of every other piece of an OS or any software running atop it.

@@

This kind of context-dependent assertion, where a fact may be true in one address space but not others, has a modal flavor. 
We propose tackling the verification of virtual memory subsystems (and kernels more broadly) by adapting ideas from hybrid
modal logic, which can label assertions true under \emph{other, named} circumstances (i.e., in another address space) with a 
modality indexed by a name for that space (in our case, the root of the page tables for an address space). This offers a 
\textit{convenient} and \textit{powerful} way to \emph{modularly}
\begin{itemize}
\item isolate assertions specific to a particular address space,
\item explicitly state when an assertion is true across address spaces,
\item manipulate address spaces from within other address spaces, and
\item reason about change in address spaces.
\end{itemize}
These advantages make this approach to reasoning about virtual memory more flexible than prior program logic techniques~\cite{kolanski08vstte,kolanski09tphols}, 
which were only able to work with a single address space (the current address space on the CPU) because they were unable
to speak directly \emph{within the logic} about other address spaces.
%\begin{figure}
%   \includegraphics[width=0.5\columnwidth]{architecture.png}
%  \caption{Layers of Implementation}
%  \todo[inline]{Colin says: I don't think this figure helps readers; it is too abstract}
%  \label{fig:architecture}
%\end{figure}

\begin{itemize}
\item We develop these ideas in the form of a logic for working with virtual-address-space-relative assertions,
      implemented as an embedded separation logic within Iris~\cite{jung2018iris} separation logic.
\item We prove the soundness of our \textsf{vProp} logic with respect to a RISC-like fragment of \textsf{AMD64} instructions.
\item We verify simplified versions of several critical virtual-memory-related pieces of OS functionality, 
      including mapping  pages, the assembly code for switching address spaces, and code which manipulates page
      tables for \emph{currently mapped memory} without disturbing mappings.\todo[inline]{Colin says: I have optimistically
      added this last example to the text here, because I think we can probably do it. Leave this TODO
      here until we either do or cut it.}
      The latter in particular goes beyond the capabilities of prior work.
\end{itemize}

### 2023-04-13 syntaxsemantics.tex
revisions up through end of section 3

\subsection{Registers and Memory}
Programs we demonstrate in this paper requires accessing two types of computer resource: registers and memory.
 A register identifier, $\reg$, is chosen from a fixed finite set of register identifiers, $\regset$. 
We use these identifiers to access the register values, $\regval \in \regvaltype$.
\begin{figure}[t]
\newcommand{\commentary}[1]{ & \text{\small\it #1} \\}
\[
  \begin{array}{r@{\;}c@{\;}l}
    \loc & \in & \Loc \\
    \reg & \in & \regset \\
    \regval & \in & \regvaltype \\
    \val & ::= & \vunit

\\
    \instrs & ::= &
    \begin{array}[t]{@{}l@{\hspace{10mm}}l@{}}
    \begin{array}[t]{@{}ll@{}}
      \iskip
                   \commentary{no-op}
      \iseq\instr\instrs
                   \commentary{sequencing}
      % \ising\instr
                   % \commentary{executing}             
    \end{array}
    %&
    %\begin{array}[t]{@{}ll@{}}
     % \ialloc\lval\allocsize
      %             \commentary{heap allocation}
    %\end{array}
    \end{array}
    \\

    \ectx & ::= &
      \hole \mid
      \iseq\ectx\instrs 
    \\
  \end{array}
\]
\caption{Syntax}
\Description{Syntax}
\label{fig:syntax}
\end{figure}
Unlike registers, we do not abstract the memory indices as a special type but instead, for the sake of clarity and ease of representation, we show differently masked machine words, $\loc \in \Loc$, with the subscripts showing the length of a 64-bit machine word after masking, e.g. $\kw{w}_{12}$ is a 12-bit resource which can be obtained after masking 52-bit of a 64-bit word.
This simple syntax in Figure \ref{fig:syntax} includes a simple syntax for values because our language is purely \textit{imperative}, i.e. an instruction does not return a value. Intentionally, we restrict any stream of instruction, $\instrs$, to eventually reduce to $\iskip$ instruction. Our syntax capture the no-op ``$\iskip$'' and the sequencing construct ``$\iseq{\instr}{\instrs}$'' in their standard forms.
\subsection{State}
\label{sec:state}
%\todo[inline]{I don't think we need or want to describe the operational semantics in detail in the submission. It adds a lot of formalization that doen'st necessarily contribute to reader understanding. We should describe the runtime \emph{state} $\sigma$ of memory and registers, since that is directly used in interpreting assertions. But the reduction rules will mostly add mental overhead here.}
%\todo[inline,color=red]{Colin says: On the contrary, we must describe assembly instructions in enough detail for readers
%who aren't already assembly programmers to decode our examples! We don't have to talk about details of the formalization
%of the semantics, but we do need to talk about basic instructions we cover, basic patterns (like destination on the left),
%and just a bit about how all memory accesses implement Figure \ref{fig:pagetables}'s semantics.}
We represent the machine state mainly as a finite map of registers to register values and a map of masked physical memory addresses to 64-bit physical memory values. Together with default \textsf{CPU} instantiation, we end up having our state with following pieces:
\begin{itemize}
\item A constant cpu initialized with default value: $\sigma.\mathcal{C}$
\item A register map: $\sigma.\mathcal{R}: \kw{greg} \rightarrow_{\textrm{fin}} \kw{regval} $
\item A memory map: $\sigma.\mathcal{M}: \Locft \rightharpoonup_{\textrm{fin}} (\Loctw \rightharpoonup_{\textrm{fin}} \Locsf )$
\end{itemize}
As one might have already anticipated from the syntax we introduce, we do not bind any value of an evaluated expression. All the indices and accessed values are treated as globally referenced. In align with this design choice, our expression is a stream of instructions, which is not evaluated to a value to be bound, but changes the machine state through the indices (e.g. $\kw{r}\in\kw{greg}$) -- to the global maps.
\subsection{Instructions}
\label{sec:instructions}
Page-table walk shown in Figure \ref{fig:pagetables} is realized with physical memory \texttt{load}s and \textsf{store}s for the entries in the page-tables which are simply modelled as lookup and updates to an instance of a memory map with 64 bit entries. In the address translation,
\begin{lstlisting}[language=Coq]
 Definition translate (rtv: word 64) (w: word 64): word 64 + MemFail :=
      l4e $\leftarrow$ translate_top_level $\sigma.\mathcal{M}$ rtv w; 
      l3e $\leftarrow$ translate_from_l4e $\sigma.\mathcal{M}$ l4e w; 
      l2e $\leftarrow$ translate_from_l3e $\sigma.\mathcal{M}$ l3e w;
      l1e $\leftarrow$ translate_from_l2e $\sigma.\mathcal{M}$ l2e w;
      addr $\leftarrow$ translate_from_l1e $\sigma.\mathcal{M}$ l1e w;
      concat (shift_dropn addr 3 three_lt64) low3.
\end{lstlisting}
we see a physical memory $\mathsf{load}$ (i.e. physical memory map ($\sigma.\mathcal{M}$) for each level) for each level of page-table-walk concluded with alignment of the returned address for each \textit{successful} page-table-walk traversal what our reasoning principles only consider.

As expected \textsf{store} would be the update on the memory map entry
\todo[inline,color=yellow]{Colin, could you mention that we simplified the maps here to get rid of MemFail?}
After defining the translation in terms of physical load and store action, now we can define our memory \lstinline|mov| instructions
\[
\begin{array}{l}
  \hbox{(\TirNameStyle{WriteToMemFromReg})} \qquad
  \textsf{mov mem64 srcreg64} \\
  \hbox{(\TirNameStyle{WriteToRegFromMem})} \qquad 
  \textsf{mov reg64 mem64}
  \end{array}
  \]
  which are realized by first translating the virtual-memory address to store a value from a register (\TirNameStyle{WriteToMemFromReg}) or load a value at the translated address to be stored in a register (\TirNameStyle{WriteToRegFromMem}). More \lstinline|mov| instructions operating with memory addresses can be found in the submission artifact for different addressing schemes and data sizes.

  Register instructions, on the other hand, are implemented as $\sigma.\mathcal{R}$ lookup and store
\[
\begin{array}{l}
  \hbox{(\TirNameStyle{WriteToRegFromReg})} \qquad 
   \textsf{mov ireg64 ireg64} \\
  \hbox{(\TirNameStyle{WriteToCtlRegFromReg})} \qquad 
  \textsf{mov creg64 ireg64}\\
    \hbox{(\TirNameStyle{WriteToRegFromCtlReg})} \qquad 
  \textsf{mov ireg64 creg64}
  \end{array}
\]
for different types of registers (e.g. control and general).
%
%\subsection{Machine State under Address Translation}
%\label{sec:selectedinstrsemantics}
%Althought we give the complete set of operational semantics rule in \sref{appendix:movops}, It is worth building the intuition on the the way some of these rules bahave in the context of address translation. In fact $\readlval\maddr{\storememstar\crval}\locsf$ is unfoled into multiple physical memory lookups for the final page address retrieval -- i.e. address translation traversal as shown in Figure \ref{fig:pagetables}:
%\begin{itemize}
%\item top-level-address translation: with the given root address (64-bit $\crval$) of the address space, performs the address translation, handling the first level (to get the PML4 entry) itself. The next level table address is computed with the fetched PML4 offset value which exhibits itself as 9-bit offset in $\kw{maddr}$. \mytodo{iso: put move}
%\item translating from PML4 entry: performs the second level of address translation, to retrieve starting at the PML4 table entry, and interprets the PML4 entry that references a Page-Directory-Pointer Table (PDPT). We obtain the PDPT offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PDT) \mytodo{iso: put move}
%\item translating from PD entry: performs the third level of address translation, to retrieve starting at the PDP table entry, and interprets a PDPTE that references a PD table. Likewise, we obtain the PD offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PT) \mytodo{iso:put move }
%\item translate from PT entry: performs final level of address translation, starting from the PT entry, with a given 12 bit page offset, we can compute the physical address referencing $\locsf$ \mytodo{ismo: put mov}
%\end{itemize}

>>>>

\subsection{Registers and Memory}
Programs we demonstrate in this paper requires accessing two types of computer resource: registers and memory.
A register identifier, $\reg$, is chosen from a fixed finite set of register identifiers, $\regset$. 
We use these identifiers $\reg$ for register names such as \texttt{rax}, \texttt{r8}, or \texttt{cr3}.
% \begin{figure}[t]
% \newcommand{\commentary}[1]{ & \text{\small\it #1} \\}
% \[
%   \begin{array}{r@{\;}c@{\;}l}
%     \loc & \in & \Loc \\
%   \end{array}
%   \begin{array}{r@{\;}c@{\;}l}
%     \reg & \in & \regset \\
%   \end{array}
%   \begin{array}{r@{\;}c@{\;}l}
%     \regval & \in & \regvaltype \\
%     % \val & ::= & \vunit
%   \end{array}
% \quad
%   \begin{array}{r@{\;}c@{\;}l}
%     \instrs & ::= &
%     \begin{array}[t]{@{}l@{\hspace{10mm}}l@{}}
%     \begin{array}[t]{@{}ll@{}}
%       \iskip
%                    \commentary{no-op}
%       \iseq\instr\instrs
%                    \commentary{sequencing}
%       % \ising\instr
%                    % \commentary{executing}             
%     \end{array}
%     %&
%     %\begin{array}[t]{@{}ll@{}}
%      % \ialloc\lval\allocsize
%       %             \commentary{heap allocation}
%     %\end{array}
%     \end{array}
%     % \\
%     % \ectx & ::= &
%     %   \hole \mid
%     %   \iseq\ectx\instrs 
%     % \\
%   \end{array}
% \]
% \caption{Syntax}
% \Description{Syntax}
% \label{fig:syntax}
% \end{figure}
For clarity and ease of representation, we use machine words, $\loc \in \Loc$, with the subscripts showing the number of bits in a word,
for memory addresses, values, and offsets, rather than distinct location types that wrap machine words.
For example, $\kw{w}_{12}$ is a 12-bit word, which can be obtained for example truncating away 52 bits of a 64-bit word ($\kw{w}_{64}$).
\subsection{State}
\label{sec:state}
We represent the machine state mainly as a finite map of registers to register values and a map of word-aligned physical memory addresses 
to 64-bit physical memory values. 
Thus our states $\sigma$ include:
%Together with default \textsf{CPU} instantiation, we end up having our state with following pieces:
\begin{itemize}
%\item A constant cpu initialized with default value: $\sigma.\mathcal{C}$
\item A register map: $\sigma.\mathcal{R}: \kw{greg} \rightarrow_{\textrm{fin}} \kw{regval} $
\item A memory map: $\sigma.\mathcal{M}: \Locft \rightharpoonup_{\textrm{fin}} (\Loctw \rightharpoonup_{\textrm{fin}} \Locsf )$
\end{itemize}
Of particular note, \texttt{cr3}, the page table register, is included in the machine state.
% As one might have already anticipated from the syntax we introduce, we do not bind any value of an evaluated expression. All the indices and accessed values are treated as globally referenced. In align with this design choice, our expression is a stream of instructions, which is not evaluated to a value to be bound, but changes the machine state through the indices (e.g. $\kw{r}\in\kw{greg}$) -- to the global maps.
\subsection{Instructions}
\label{sec:instructions}

\begin{figure}[t]
\begin{lstlisting}[language=Coq,mathescape]
 Definition translate (rtv: word 64) ($\sigma$: state) (w: word 64): word 64 + MemFail :=
      l4e $\leftarrow$ translate_top_level $\sigma.\mathcal{M}$ rtv w; 
      l3e $\leftarrow$ translate_from_l4e $\sigma.\mathcal{M}$ l4e w; 
      l2e $\leftarrow$ translate_from_l3e $\sigma.\mathcal{M}$ l3e w;
      l1e $\leftarrow$ translate_from_l2e $\sigma.\mathcal{M}$ l2e w;
      addr $\leftarrow$ translate_from_l1e $\sigma.\mathcal{M}$ l1e w;
      return concat (shift_dropn addr 3 three_lt64) low3.
\end{lstlisting}
\vspace{-1em}
\caption{\coq implementation of address translation (slightly simplified).}
\label{fig:coq_addr_translation}
\end{figure}

Programs in our logic are instruction sequences \instrs, which are formed by either a basic instruction \iskip, or prefixing an existing instruction
sequence with an additional instruction (\iseq\instr\instrs).
We model (and later, give program logic rules for) instructions for basic register moves, and reading and writing memory.
The latter require page table walks.
Figure \ref{fig:coq_addr_translation} gives a slightly simplified version of our address translation code in monadic style:
starting from the root page table address \textsf{rtv}, the virtual address \textsf{w} is either resolved to a physical address or an error is returned.
The full set of possible errors is not relevant to this paper; our logic is tailored for kernel code which should not fault,
so our proof rules guarantee no memory failures occur --- that the page table walk for any dereferenced address will succeed.

% Page-table walk shown in Figure \ref{fig:pagetables} is realized with physical memory \texttt{load}s and \textsf{store}s for the entries in the 
% page-tables which are simply modelled as lookup and updates to an instance of a memory map with 64 bit entries. In the address translation,
% we see a physical memory $\mathsf{load}$ (i.e. physical memory map ($\sigma.\mathcal{M}$) for each level) for each level of page-table-walk 
% concluded with alignment of the returned address for each \textit{successful} page-table-walk traversal which is what our reasoning principles 
% only consider.

The main instructions we implement are variants of the x86-64 \lstinline|mov| instruction, which we format in Intel syntax (destination on the left, source on the right).
As expected \textsf{store} would be the update on the memory map entry
% After defining the translation in terms of physical load and store action, now we can define our memory \lstinline|mov| instructions
For memory operations, we treat
\[
\begin{array}{l}
  \hbox{(\TirNameStyle{WriteToMemFromReg})} \qquad
  \textsf{mov}~[r_m]~r_r \\
  \hbox{(\TirNameStyle{WriteToRegFromMem})} \qquad 
  \textsf{mov}~r_r~[r_m]
  \end{array}
  \]
which are realized by first translating the virtual-memory address of a memory location stored in $r_m$ to a physical memory location,
then updating those memory contents with the contents of register $r_r$ (\TirNameStyle{WriteToMemFromReg}) 
or loading the value at that physical memory location into the register $r_r$
(\TirNameStyle{WriteToRegFromMem}). 
% More \lstinline|mov| instructions operating with memory addresses can be found in the submission artifact for different addressing schemes and data sizes.
Additional variants (such as instructions which access memory at a constant offset from the virtual address stored in a register) are also treated.

For register operations, we support moves back and forth between any pair of registers, though the \coq formalization splits these according to whether
the source or destination registers are standard data registers (e.g., \lstinline|rax|) or control registers (e.g., \lstinline|cr3|).
% \[
% \begin{array}{l}
%   \hbox{(\TirNameStyle{WriteToRegFromReg})} \qquad 
%    \textsf{mov ireg64 ireg64} \\
%   \hbox{(\TirNameStyle{WriteToCtlRegFromReg})} \qquad 
%   \textsf{mov creg64 ireg64}\\
%     \hbox{(\TirNameStyle{WriteToRegFromCtlReg})} \qquad 
%   \textsf{mov ireg64 creg64}
%   \end{array}
% \]
% for different types of registers (e.g. control and general).
%
%\subsection{Machine State under Address Translation}
%\label{sec:selectedinstrsemantics}
%Althought we give the complete set of operational semantics rule in \sref{appendix:movops}, It is worth building the intuition on the the way some of these rules bahave in the context of address translation. In fact $\readlval\maddr{\storememstar\crval}\locsf$ is unfoled into multiple physical memory lookups for the final page address retrieval -- i.e. address translation traversal as shown in Figure \ref{fig:pagetables}:
%\begin{itemize}
%\item top-level-address translation: with the given root address (64-bit $\crval$) of the address space, performs the address translation, handling the first level (to get the PML4 entry) itself. The next level table address is computed with the fetched PML4 offset value which exhibits itself as 9-bit offset in $\kw{maddr}$. \mytodo{iso: put move}
%\item translating from PML4 entry: performs the second level of address translation, to retrieve starting at the PML4 table entry, and interprets the PML4 entry that references a Page-Directory-Pointer Table (PDPT). We obtain the PDPT offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PDT) \mytodo{iso: put move}
%\item translating from PD entry: performs the third level of address translation, to retrieve starting at the PDP table entry, and interprets a PDPTE that references a PD table. Likewise, we obtain the PD offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PT) \mytodo{iso:put move }
%\item translate from PT entry: performs final level of address translation, starting from the PT entry, with a given 12 bit page offset, we can compute the physical address referencing $\locsf$ \mytodo{ismo: put mov}
%\end{itemize}
